---
title: "Social media in comparative politics"
author: "Sebastian Stier"
lesson: 6
institute: Teaching Presentation University of Hamburg, Summer semester 2025
date: "2025-05-13"
date-format: "YYYY-MM-DD"
bibliography: references.bib
footer: "[Data Science in Comparative Politics](https://sebstier.github.io/ds_comp_pol25/)"
format: 
  fakegesis-revealjs: 
    code-line-numbers: false

---


##

```{r, echo=FALSE, warning=FALSE, out.width="105%", message=FALSE}
library(openxlsx)
library(tidyverse)
library(gapminder)
library(plotly)
library(gt)
library(gtExtras)
read.xlsx("../material/data/schedule.xlsx") %>%
    rename(`Required reading` = "Required.reading") %>%
    head(6) %>% 
    gt() %>%
    tab_header(md("**Seminar dates and topics**")) %>%
    #tab_header("**Seminar dates and topics**") %>%
    #fmt_markdown() %>% #columns = TRUE
    # cols_width(Date ~ px(400)#,
    #            #Topics ~ px(350)#,
    #            #`Required reading` ~ px(400)
    #            ) %>%
    cols_width(Date ~ pct(20),
               Topics ~ pct(30),
               `Required reading` ~ pct(50)
               ) %>%
    tab_options(data_row.padding = px(3)) %>%
    tab_options(heading.title.font.size = 14,
                column_labels.font.weight = "bold",
                table.font.size = 13.5) %>%
     gt_highlight_rows(
     rows = 6,
     fill = "lightblue"
   ) %>% 
  fmt_markdown() %>% 
  cols_align(
    align = "left",
    columns = everything()
  )


```


## Agenda for today


1. Recap

2. Social media data in political science

3. Text as data in *R*

4. Text as data in political science


## Learning objectives for today

After this class you will be able to... 

1. describe the role of social media and politics,
2. critically reflect upon the value of social media data in comparative politics,
3. work with text as data using *quanteda* in R.


# 1. Recap {background-color="#58748F"}


```{r}
# Inspecting the dataset
head(gapminder)

```


```{r}
#| echo: fenced
#| output-location: column-fragment
#| code-line-numbers: "|2"

library(ggplot2)

mtcars |> 
  ggplot(aes(x = disp, y = mpg)) +
  geom_point() +
  geom_smooth(method = "loess", formula = "y~x")
```


# 2. Social media data in political science {background-color="#58748F"}


## Think-pair-share

1. Discuss the literature for today with your neighbor (5 min.)
2. Focus on the used methods, their pros and cons, feasibility, etc. 
3. Locate our two required readings on the map of @grimmer_text_2013
4. Share and discuss your results with the full class



## Format of the reviews
- Guiding questions:

::: {style="font-size: 90%;"}
    - Describe what problem or question these papers address and the main contributions that they make towards a solution or answer. 
    
    - Reflect on the main strengths and weaknesses of these papers.
    
    - What are the concepts, assumed social mechanisms and operationalizations used in the papers?
    
    - What are the data in use? How are the data analyzed? Does this seem appropriate?
    
    - How convincing do you find the arguments presented by the authors? Are the conclusions backed up by the empirical results?
    
    - Do you see any ethical concerns?
    
    - If papers are empirical, how do you judge their replicability?
    
    - If applicable, are there commonalities or differences between the two papers? How do they relate to each other?
    
:::



# 3. Text as data in *R* {background-color="#58748F"}


## Automated text analysis in *R*

![](../material/img/quanteda){fig-align="center"}

::: {style="font-size: 30%;"}
Benoit, K., Watanabe, K., Wang, H., Nulty, P., Obeng, A., Müller, S., Matsuo, A., Perry, P. O., Kuha, J., & Lauderdale, B. (2018). quanteda: An R package for the quantitative analysis of textual data. [*Journal of Open Source Software*, 3(30), 774.](https://joss.theoj.org/papers/10.21105/joss.00774)

:::


## Trump Twitter Archive


![](../material/img/trump_twitter){width=100%}


::: {style="font-size: 40%;"}
[https://www.thetrumparchive.com](https://www.thetrumparchive.com)
:::


## Coding
- Download file **06_text_as_data.R** from [https://sebastianstier.com/ds_comp_pol25/material.html](https://sebastianstier.com/ds_comp_pol25/material.html)
- Open it in *RStudio* within your project



# 4. Text as data in political science {background-color="#58748F"} 


## Automated text analysis: The menu of options

![](../material/img/grimmer_stewart){fig-align="center"}

::: {style="font-size: 30%;"}
Grimmer, J., & Stewart, B. M. (2013). Text as Data: The Promise and Pitfalls of Automatic Content Analysis Methods for Political Texts. [*Political Analysis*, 21(3), 267--297.](https://doi.org/10.1093/pan/mps028)

:::



# Thank you for your attention! See you on May 20 2025 {background-color="#58748F"} 

## References
