---
title: "Research Ethics and Web Data"
author: "Sebastian Stier"
lesson: 2
institute: University of Mannheim & GESIS -- Leibniz Institute for the Social Sciences
date: "2024-02-28"
date-format: "YYYY-MM-DD"
bibliography: references.bib
footer: "[Computational Social Science and Digital Behavioral Data, University of Mannheim](https://sebastianstier.com/ma_css24/)"
format: 
  fakegesis-revealjs: 
    code-line-numbers: false

---


## Agenda for today

1. Recap

2. Warmup session in *R*

3. Research ethics and web data

4. *R*: First exploration of digital behavioral data




# 1. Recap {background-color="#58748F"} 



## What are digital behavioral data?

. . . 

:::: {.columns}

::: {.column width="55%"}
![](../material/img/platform_pics){.figure width="710"}

:::

::: {.column width="40%"}

::: {style="font-size: 85%;"}

"Records of activity (trace data) undertaken through an online information system (thus, digital)" (Howison et al., 2011: 769) that can be collected from a multitude of technical systems, such as websites, social media platforms, smartphone apps, or sensors.


:::

::: {style="font-size: 40%;"}
Howison, J., Wiggins, A., & Crowston, K. (2011). Validity issues in the use of social network analysis with digital trace data. [*Journal of the Association for Information Systems*, 12(12), 767–797, 769.](https://aisel.aisnet.org/cgi/viewcontent.cgi?article=1594&context=jais)

Stier, S., Breuer, J., Siegers, P., & Thorson, K. (2020). Integrating Survey Data and Digital Trace Data: Key Issues in Developing an Emerging Field. [*Social Science Computer Review*, 38(5), 503–516, 504.](https://journals.sagepub.com/doi/epub/10.1177/0894439319843669)


:::

:::

::::


## What is computational social science?

. . . 

Computational social science is an interdisciplinary field that advances theories of human behavior by applying computational techniques to large datasets from social media sites, the Internet, or other digitized archives such as administrative records. Our definition forefronts sociological theory because we believe the future of the field within sociology depends not only on novel data sources and methods, but also on its capacity to produce new theories of human behavior or elaborate on existing explanations of the social world.

::: {style="font-size: 50%;"}
Edelmann, A., Wolff, T., Montagne, D., & Bail, C. A. (2020). Computational Social Science and Sociology. [*Annual Review of Sociology*, 46(1), 61–81: 62.](https://doi.org/10.1146/annurev-soc-121919-054621)


:::


## The role of theory and existing knowledge

"Our definition forefronts sociological theory because we believe the future of the field within sociology depends not only on novel data sources and methods, but also on its capacity to produce new theories of human behavior or elaborate on existing explanations of the social world." [@edelmann_computational_2020: 62]


## Explanation vs. description I

:::: {.columns}

::: {.column width="40%"}
![](../material/img/guess_quant_descr){width=80%}

::: {style="font-size: 26%;"}

Munger, K., Guess, A. M., & Hargittai, E. (2021). Quantitative Description of Digital Media: A Modest Proposal to Disrupt Academic Publishing. [*Journal of Quantitative Description: Digital Media*, 1.](https://doi.org/10.51685/jqd.2021.000)


:::

:::

::: {.column width="60%"}

::: {style="font-size: 72%;"}


"First and foremost, we respond to an undersupply of quantitative descriptive research in social science. Causal research that asks the question why has largely taken the place of descriptive research that asks the question what. Gerring (2012) diagnosed a general tendency to dismiss 'Mere Description' as a 'mundane task ... of little intrinsic scientific value,' advocating instead that it be taken seriously as part of the general social scientific method. 

We firmly agree. However, critique alone does not change the material conditions and incentives of practicing academics; we see this journal as a practical step towards raising the status of description as a method." [@munger_quantitative_2021: 3-4]


:::

:::

::::


## Explanation vs. description II

:::: {.columns}

::: {.column width="40%"}
![](../material/img/guess_quant_descr){width=80%}

::: {style="font-size: 26%;"}
Munger, K., Guess, A. M., & Hargittai, E. (2021). Quantitative Description of Digital Media: A Modest Proposal to Disrupt Academic Publishing. [*Journal of Quantitative Description: Digital Media*, 1.](https://doi.org/10.51685/jqd.2021.000)

:::

:::

::: {.column width="60%"}

::: {style="font-size: 72%;"}

Second, rather than define our new venue in terms of existing disciplinary boundaries, we instead embrace a topical focus on digital media, broadly construed. We argue that the centrality and dynamism of digital media — information and communication technologies, including social media, that increasingly structure the way people interact with the world — necessitates increased scholarly energy devoted to sustained, continuous, quantitative description. [...] 

Today, there are more hours of video uploaded to YouTube every day than were broadcast in the 1950s U.S. in a year. The daily content of Twitter is different (and different in unpredictable ways) than it was the day before." [@munger_quantitative_2021: 3-4]


:::

:::

::::


:::: {.incremental}

## Examples of novel sociotechnical phenomena

. . .

- Harmful online communication
  - E.g., misinformation and deception
  - E.g., violent language, harassment, racism
- Algorithmic influence, bias and discrimination
- Filter bubbles and echo chambers
- Social bots 
- Virality of online content and memes
- Online inequalities

::::

##

```{r, echo=FALSE, warning=FALSE, out.width="105%", message=FALSE}
library(openxlsx)
library(tidyverse)
library(gt)
library(gtExtras)
read.xlsx("../material/data/schedule.xlsx") %>%
    rename(`Required reading` = "Required.reading") %>%
    gt() %>%
    tab_header(md("**Seminar dates and topics**")) %>%
    fmt_markdown(columns = TRUE) %>%
    cols_width(Date ~ px(200),
               Topics ~ px(250),
               Topics ~ px(500)) %>%
    tab_options(data_row.padding = px(5.4)) %>%
    tab_options(heading.title.font.size = 14,
                table.font.size = 12) %>% 
     gt_highlight_rows(
     rows = 2,
     fill = "lightblue"
   )

```


# 2. Warmup session in R {background-color="#58748F"} 


## Getting back to coding

Do you have any *R*-related questions?

. . . 

**Styleguide**
Guidance for coding in *R*: [https://style.tidyverse.org/index.html](https://style.tidyverse.org/index.html)

```{r, eval=FALSE,echo=FALSE,message=FALSE}
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```


# 3. Research ethics and web data {background-color="#58748F"}


## Research ethics?

:::{.incremental}

*Research ethics*

- Moral principles and actions guiding and shaping research

    - from inception to completion,
    - through dissemination and sharing of findings,
    - including archiving and future use.
:::

. . .

:::{.incremental}

*Research ethics in the social sciences*

- Initially "patient protection" model of medical research
- Today broader scope including consideration of benefits, risks and harms to all persons connected with the research
- Including social responsibilities of researchers

:::


## Who are the relevant actors?

. . . 

![](../material/img/actors_data_protection){width=100%}


## Data protection


- Data protection
    - Part of fundamental right to privacy (or "informational freedom")

. . . 

- "Privacy is a personal condition of life characterised by seclusion from, and therefore absence of acquaintance by, the public" [@neethling2005personality: 233]
- Prevention of unwanted disclosure of personal information or the misuse of such information


## Data protection: legal foundations

- Prevention of unwanted disclosure of personal information or the misuse of such information
    - Core of data protection
- Legal framework in the European Union
    - Charter of Fundamental Rights of the EU (Art. 8)
    - General Data Protection Regulation (GDPR)
    - National and sub-national data protection acts
    - Specialized laws


## Informed consent

![](../material/img/informed_consent){fig-align="center"}

- Bridge between data protection and research ethics

. . .

- Informed consent means for example:
    - information
    - transparency
    - minimal requirement $\rightarrow$ chance *not* to consent
- Status quo in social media research: lack of informed consent


## Back to coding
- Go back to file **2_session_two.R** in [https://sebastianstier.com/ma_css24/material.html](https://sebastianstier.com/ma_css24/material.html)


# Thank you for your attention! See you on March 13th {background-color="#58748F"} 

## References
